---
title: "final project Trim28"
author: "Shaqed Carasso"
date: "2/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project overview

placeholder

```{r message=FALSE, warning=FALSE, include=TRUE, paged.print=FALSE}

library(dplyr)
library(ggplot2)
library(MetBrewer)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(ggpubr)
library(limma)
library(reshape2)
library(RColorBrewer)
library(corrplot)
library("org.Mm.eg.db")
```

## Open files, create counts and metadata tables, visualize library sizes

open files

```{r}

counts_file <- "trim28_project/E-MTAB-9809-raw-counts.tsv"
coldata_file <- "trim28_project/E-MTAB-9809-experiment-design.tsv"

counts <- read.table(counts_file, sep = '\t',header = T, row.names = 1, stringsAsFactors = TRUE)
metadata <-  read.table(coldata_file, row.names = 1, header = T, sep = '\t', 
                        stringsAsFactors = TRUE)
```

keep a 'dictionary of gene ids and names, then remove gene names from counts file

```{r}
geneNames <- as.data.frame(counts[,"Gene.Name"])
geneNames$Gene.ID <- rownames(counts)
geneNames <- geneNames %>% rename('counts[, "Gene.Name"]'= "Gene.Name")
counts <- subset( counts, select = -Gene.Name)
```

get only informative columns of metadata, sex and group

```{r}
metadata <- metadata %>% select('Sample.Characteristic.genotype.', 'Sample.Characteristic.sex.') %>%
        rename('Sample.Characteristic.sex.' = "Sex") %>%
        mutate(Group = ifelse(grepl("Trim28", Sample.Characteristic.genotype.), "Trim28KO", "WT"))


```

Library sizes:

```{r}
metadata$lib_size = colSums(counts)

ggplot(metadata, aes(fill = Group)) +
        geom_col(aes(x = rownames(metadata), y = lib_size)) +
        labs(title = "Barplot of library sizes") +
        #scale_x_discrete(limits = rownames(metadata)) +
        theme_pubr() +
        xlab(NULL) +
        ylab("Library size") +
        scale_fill_manual(values=met.brewer("Isfahan1", 2)) +
        facet_wrap(~Sex, scales = "free") +
        theme(axis.text.x = element_text(angle=90, vjust=0.4))

```


Normalize to CPM, exclude lowly expressed genes:

```{r}
cpm <- apply(counts, 2, 
             function(x) as.numeric(x)/sum(as.numeric(x)) * 10^6)

colSums(cpm)

keep <- rowMeans(cpm) > 0.3
table(keep)

counts.filtered = counts[keep,]
dim(counts.filtered)

```

Quality control, looking at correlations between the samples and PCA:

```{r}
# Compute the correlation values between samples
correlationMatrix <- cor(cpm) 

breaksList <- seq(0.5, 1, by = 0.001)
heatmap.corr <- pheatmap(correlationMatrix, annotation = select(metadata, c(Group, Sex)), 
                         #annotation_colors = list(Group = c('none' = "lightblue3", immunized = "red3", 
                          #                                  "immunized_cross" = "salmon")),
                         breaks = breaksList,
                         # cluster_rows = FALSE,
                         #cluster_cols = FALSE,
                         
                         color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaksList)),
                         main = "Correlations Heatmap")

```

ERR4873315 is very different from the rest, so I think it's better to filter it.

Create DeseQ object, comapring Groups with Sex as a confounding factor:


```{r}
metadata_filtered_samples <- subset(metadata, row.names(metadata) != ("ERR4873315"))
subset_Samples <- as.vector(rownames(metadata_filtered_samples))
subset_reads <- counts.filtered[, subset_Samples]

countdata.deseq <- DESeq2::DESeqDataSetFromMatrix(countData = subset_reads, 
                                                  colData = metadata_filtered_samples, 
                                                  design = ~ Group + Sex)


dds <- DESeq(countdata.deseq)
# Extract the normalized counts
countdata_normalized_counts <- counts(dds, normalized = TRUE)
```
Dispersion (why?)

```{r}
plotDispEsts(dds)
```

Plotting a PCA plot shows that the samples are divided by sex across PC1, and by their condition on PC2
```{r}
# Plot PCA 

# Extracting transformed values, selecting top 1000 most variable genes

vsd <- vst(dds, blind=FALSE, nsub = 1000)


pcaData <- plotPCA(vsd, intgroup = c("Group", "Sex"), returnData = TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"),digits = 1)

ggplot(pcaData, aes(x = PC1, y = PC2, color = Group)) +
        coord_fixed()+
        geom_point(size = 2, aes(shape = Sex)) +
        xlab(paste0("PC1: ", percentVar[1], "%")) + 
        ylab(paste0("PC2: ", percentVar[2], "%")) +
        #geom_text(label = rownames(pcaData), size = 4, nudge_y = 0.2, show.legend = FALSE)+
        stat_ellipse(geom = "polygon", type="norm", alpha=0.1, aes(fill=Group)) +
        labs(color = "Group") +
        guides(color = guide_legend(order = 1),
               shape = guide_legend(order = 2))+
        scale_fill_manual(values=met.brewer("Isfahan1", 2), guide = NULL) +
        scale_color_manual(values=met.brewer("Isfahan1", 2)) +
        theme_pubr(legend = "right")
        
```
deseq results

```{r}

res <- results(dds, 
               contrast = c("Group", "Trim28KO", 'WT'), 
               alpha = 0.05, 
               lfcThreshold = 0)



summary(res)


hist(res$pvalue[res$baseMean > 1], breaks = 0:20/20,
     col = "grey50", border = "white")



resultsNames(dds)
```

```{r}
resLFC <- lfcShrink(dds, coef="Group_WT_vs_Trim28KO", type="apeglm")


resLFC$entrez <- mapIds(org.Mm.eg.db,
                     keys=rownames(resLFC),
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")




dds_res_all <- data.frame(res)

dds_res_all_named <- left_join(rownames_to_column(dds_res_all), geneNames, by=c("rowname" = "Gene.ID"))


ggplot() +
        geom_point(data = dds_res_all, mapping = aes(x = log2FoldChange, y = -log10(pvalue)), 
                   col = "grey80", pch = 21, size = 4) + 
        geom_point(data = subset(dds_res_all, log2FoldChange < -0.58 & padj < 0.05),
                   aes(log2FoldChange, -log10(pvalue)), color = "red", size = 4)  +
        geom_point(data = subset(dds_res_all, log2FoldChange > 0.58 & padj < 0.05),
                   aes(log2FoldChange, -log10(pvalue)), color = "green3", size = 4) +
        #ggtitle("in-vitro -CD3 (red) vs none (green) \nsignificant genes are colored") +
        geom_text(data = subset(dds_res_all, abs(log2FoldChange) > 1 & padj < 0.05),
                  aes(log2FoldChange, -log10(pvalue), label = entrez), 
                  vjust =1, size = 4)+
        scale_color_manual(values=met.brewer("Isfahan1", 2)) +

        theme_pubr()

dds_res_all <- data.frame(res) %>% mutate(threshold = (padj < 0.05 & abs(log2FoldChange) >= 0.58)) %>%
        filter(!is.na(padj))


dds_res_all <- left_join(rownames_to_column(dds_res_all), geneNames, by=c("rowname" = "Gene.ID"))


ggplot(dds_res_all) + 
        geom_point(aes(x = log2FoldChange, y = -log10(pvalue), color = threshold)) + 
        xlab("log2 fold change") + 
        ylab("-log10 adjusted p-value") + 
        theme(legend.position = "none", 
              plot.title = element_text(size = rel(1.5), hjust = 0.5), 
              axis.title = element_text(size = rel(1.25)))+
        scale_color_manual(values=met.brewer("Isfahan1", 2)) +
        #geom_text(data = subset(dds_res_all, abs(log2FoldChange) > 1 & padj < 0.05),
        #          aes(log2FoldChange, -log10(pvalue), label = entrez), 
         #         vjust =1, size = 4)+
        
        
        geom_text(data = subset(dds_res_all, Gene.Name %in% c("Elovl3", "Olr1", " Elovl7", " Klf14")),
                  aes(log2FoldChange, -log10(pvalue), label = Gene.Name), 
                  vjust =1, size = 4)
        theme_pubr(legend = "none")


```


enrichment


glm\rforest





